The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
edits on split thm chapter

\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}

@@

@@

When analysing the complexity of the $\rnni$ space, the comparison to $\nni$ is an important step.
The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.

\begin{theorem}
	There are trees $T,R$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T$ to $R$.
	\label{thm:split_nni}
\end{theorem}

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
Let $T$ and $R$ be two caterpillar trees.
If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T$ and $R$, the resulting trees $R$ and $R'$ have distance $d_{\nni}(R,R') = d_{\nni}(T,R)$ in the $\nni$ graph.
When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $R$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $R$ is tranformed to a tree $R'$ that contains the same cherries as $R$.
Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are some rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

Therefore, the following conjecture was raised~\cite{Gavryushkin2017}.

@@

However, we found a counterexample to this conjecture:

>>>>

\newcommand{\np}{\mathcal{NP}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\rspr}{\mathrm{rSPR}}

@@

\todo{define splits and delete definition from split thm chapter!}

@@

%Shall we put this bit into the introduction?
It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$~\cite{jiang2000}, $\rspr$~\cite{Bordewich2005}, or $\tbr$~\cite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.

The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.

\begin{theorem}
	There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
	\label{thm:split_nni}
\end{theorem}

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The clusters given by the two edges incident to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

Since the example given in~\cite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised~\cite{Gavryushkin2017}.

@@

Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
However, there is a simple counterexample to this conjecture:
